---
title: "Predicting house prices using Linear Regression in R"
author: "Eduardo Marquez Mora"
date: "2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load package
library(MASS)
library(car)

# Load dataset
data(Boston)

attach(Boston)

```

## Introduction

Using the built-in Boston dataset, predict house prices with information such as information such as:

* Crime rate
* Average number of rooms
* Property tax rate
* Distance to employment centers
* Median home value (medv)


```{r Boston}

head(Boston)
```

## 1. How many observations and variables are in the dataset?

```{r 1}
dim(Boston)
```
501 observations and 14 variables.

## 2. What is the average house price?

```{r 2}
mean(Boston$medv)
```
Around $22,533.

## 3. Which variable has the strongest correlation with house price?

```{r 3}

cor_matrix_Boston <- cor(Boston)
round(cor_matrix_Boston, 3)

```
rad (index of accessibility to radial highways) and tax (full-value property0tax rate per $10,000) seem to have the highest correlation: 0.910

## 4. Create a scatterplot of:

* rm (average rooms)
* versus medv (median value of home)

```{r 4}

plot(Boston$rm, Boston$medv,
     xlab = "Average Rooms",
     ylab = "Median Home Value",
     main = "Rooms vs House Price")

```


## 5. Fit a simple linear regression model predicting house price (medv) using number of rooms (rm)

```{r 5}

simpleBostonModel <- lm(medv ~ rm, data = Boston)
summary(simpleBostonModel)

```
Model: y = -34.671 + 9.102x + E


With the given model, we can predict that the more rooms, the higher the house price will be. More specifically, the house price will increase by approximately $9,100, holding all other variables constant.

## 7. What percentage of variation in house price is explained by the model?

With the previous output, we can observe R-squared: 0.4835. About 48.35% of the variation in median house prices is explained by the number of rooms in the house.

## 8. Multiple linear regression. Fit a multiple linear regression model using rm, lstat, ptratio to predict medv

```{r 8}

multipleBostonModel <- lm(medv ~ rm + lstat + ptratio, data = Boston)
summary(multipleBostonModel)

```
Model: y = 18.567 + 4.515x1 - 0.572x2 - 0.931x3 + E

* Each additional room is associated with a increase of $4,515 dollars in median home value.
* For a 1-unit increase in lstat (lower status of the population percentage), the median home value is expected to decrease by $572.
* For a 1-unit increase in ptratio (pupil-teacher ratio by town),, the median home value is expected to decrease by $931.

## 9. Which predictors are statistically significant?

All of the predictors are statistically significant, since their p-values are very much smaller than 0.05.

## 10. Compare both models

```{r 10}

summary(simpleBostonModel)
summary(multipleBostonModel)

```

The multiple regression model has a higher Adjusted R-squared, as well as the addition of other significant predictors.


## 11. Check residual plots for multiple regression model

```{r 11}
par(mfrow = c(2,2))
plot(multipleBostonModel)

```

There is a clear violation in the linearity and constant variance assumptions.

## 12. Create confidence interval for regression coefficients

```{r 12}
confint(multipleBostonModel)
```


## 13. Check for multicollinearity using VIF

```{r 13}
vif(multipleBostonModel)
```
The VIF for all variables is under our threshold of 10, so there is no risk of multicollinearity.


## 14. House price prediction 
House: rm = 7, lstat = 10, ptratio = 18

```{r 14}
newHousePrediction <- data.frame(
  rm = 7,
  lstat = 10,
  ptratio = 18
)

predict(multipleBostonModel, newdata = newHousePrediction)
```
The predicted house value is $27,704.

